---
title: "Limitations and unimplemented behaviour"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Limitations and unimplemented behaviour}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_desolve <- requireNamespace("deSolve", quietly = TRUE)
```

This vignette is the honest inventory: what `r2sbml` does not do, what it does
only partially, and — the part that matters most in practice — **which of those
gaps announce themselves and which do not**.

Read `vignette("r2sbml", package = "r2sbml")` first for how the package is meant
to be used.

```{r library}
library(r2sbml)
ex <- function(f) system.file("examples", f, package = "r2sbml")
```

# The single most important table

Generated code that fails loudly is a nuisance. Generated code that runs and
quietly returns the wrong numbers is a hazard. Here is where each limitation
falls:

| Limitation | How you find out |
| --- | --- |
| Algebraic rules in mrgsolve / rxode2 / ubiquity | **R warning** + comment in the file |
| Algebraic rules that do not form a square system | **R warning** + comment |
| Dilution for an assignment-rule compartment | **R warning** |
| Constructs with no ubiquity spelling (incl. `delay`) | **R warning** + `# WARNING` block in the file |
| `delay` in the other five targets | **silent** — code fails at run time |
| Events | **silent** — a count in the header comment only |
| Two-argument `log(b, x)` in R and MATLAB | **silent** — wrong answer |
| `root`, `piecewise`, `factorial`, `%` in some targets | **silent** — untested |
| Inconsistent initial conditions for a DAE | solver error, not mentioning SBML |
| A semantically invalid model | **silent** — nothing is validated |

Everything below expands on these.

# Scope: reading and exporting only

`r2sbml` has no setter API. You cannot change a species' initial value, add a
reaction, or write SBML back out. There is no `setSpecies()`, no
`writeSBML()`, and `getModel()` returns an opaque external pointer rather than
an R-side object you could edit.

If you need to modify models programmatically, use libSBML's own R bindings or
edit the XML directly. `r2sbml` is for inspecting a model and getting simulation
code out of it.

# Per-target capability matrix

| Feature | deSolve | mrgsolve | rxode2 | MATLAB | Julia | ubiquity |
| --- | :-: | :-: | :-: | :-: | :-: | :-: |
| Rate rules / reactions | yes | yes | yes | yes | yes | yes |
| Assignment rules | yes | yes | yes | yes | yes | yes |
| Boundary / constant species | yes | yes | yes | yes | yes | yes |
| Initial-condition unit conversion | yes | yes | yes | yes | yes | yes |
| Algebraic rules (as a DAE) | yes | no | no | yes | yes | no |
| Compartment with a rate rule | yes | yes | yes | yes | yes | yes |
| Compartment dilution term | yes | yes | yes | yes | yes | yes |
| Compartment with an assignment rule | partial | partial | partial | partial | partial | partial |
| `delay` | no | no | no | no | no | no |
| Events | no | no | no | no | no | no |
| Exponentiation spelling | `^` | `pow()` | `^` | `^` | `^` | `SIMINT_POWER` |

"partial" for an assignment-rule compartment means the volume itself is correct
but the dilution term is missing; see below.

# Algebraic rules

An algebraic rule is a constraint `0 = f(...)`, not a derivative, so a model
carrying one is a differential-algebraic system rather than an ODE.

## Where a DAE is emitted

deSolve, MATLAB and Julia can each solve one, and `r2sbml` emits a real DAE for
them — a `daspk` residual function, or a singular mass matrix. That path is
demonstrated in the main vignette. The constraint is genuinely enforced:

```{r dae-works, eval = has_desolve}
out <- tempfile(fileext = ".R")
convertReactions(ex("sbmlalgebraicrules.xml"), out, format = "R")

env <- new.env(); source(out, local = env)
d <- as.data.frame(deSolve::daspk(y = env$InitialAmounts, dy = env$InitialDerivatives,
                                  times = seq(0, 10, by = 2), res = env$massBalances,
                                  parms = env$parameters))
max(abs(d$E + d$ES - d$E_total))    # constraint residual, ~1e-13
```

## Where it is not

mrgsolve, rxode2 and ubiquity integrate ODEs only. There the rule is written out
as a comment, the species it constrains is left at a zero derivative, and a
warning is raised:

```{r dae-warns}
tryCatch(
  invisible(capture.output(convertReactions(ex("sbmlalgebraicrules.xml"),
                                            tempfile(), format = "mrgsolve"))),
  warning = function(w) cat(conditionMessage(w), "\n")
)
```

There is no fix short of solving the constraint symbolically for its variable
and emitting that as an assignment, which only works where such a solution
exists and is unique.

## The square-system requirement

SBML does not record *which* variable a given algebraic rule determines, and
libSBML exposes no matching for it. `r2sbml` works out the candidate set
instead — a species that is not constant and carries neither a rate rule nor an
assignment rule is undetermined, and the specification requires the algebraic
rules to determine exactly those.

A DAE is emitted only when those two counts agree. If they differ, the model is
over- or under-determined as far as this heuristic can tell, and rather than
guess, `r2sbml` falls back to comments and warns. A proper bipartite matching
over the variables each rule references would widen the set of models that
qualify; that is not implemented.

## Consistent initial conditions are assumed, not checked

A DAE has to start on the constraint manifold. `sbmlalgebraicrules.xml` happens
to. A model that does not will fail inside the solver — `daspk` cannot find
consistent initial conditions, `ode15s` rejects the initial state — and **the
error will not mention SBML or `r2sbml`**, which makes it hard to trace back to
the model.

`daspk`'s `estini` argument can estimate consistent values, but it requires the
algebraic equations to come last in the residual vector, which the current row
ordering does not guarantee.

# `delay`

SBML's `delay` csymbol passes through unchanged into every target:

```{r delay}
out <- tempfile(fileext = ".R")
invisible(capture.output(convertReactions(ex("sbmldelay.xml"), out, format = "R")))
grep("delay", readLines(out), value = TRUE)
```

`delay()` is not a function in base R, in C++, in MATLAB or in Julia, so this
code does not run anywhere. A delay-differential equation needs a different
solver interface entirely — `deSolve::lagvalue()` for R, and the equivalent per
target.

Only the ubiquity writer notices:

```{r delay-ubiquity}
tryCatch(
  invisible(capture.output(convertReactions(ex("sbmldelay.xml"),
                                            tempfile(), format = "ubiquity"))),
  warning = function(w) cat(conditionMessage(w), "\n")
)
```

For the other five targets this is **silent**. `sbmldelay.xml` is the one
example model whose generated deSolve, mrgsolve, rxode2, MATLAB and Julia output
will not run.

# Events

Events are not exported at all. They survive as a number in the header comment
and nothing else:

```{r events}
out <- tempfile(fileext = ".R")
invisible(capture.output(convertReactions(ex("sbmlevent.xml"), out, format = "R")))
grep("events", readLines(out), value = TRUE)
```

`sbmlevent.xml` has two events, and neither appears in the generated code. The
simulation will run — and will silently ignore every trigger, delay and event
assignment in the model.

You can still inspect them from R:

```{r events-inspect}
getEventMath(getModel(ex("sbmlevent.xml")))
```

Translating them would mean mapping SBML triggers onto each target's own
mechanism — deSolve's `events`/`roots`, mrgsolve's `$EVENT`, and so on — which
differ enough that there is no shared representation to generate from.

# Compartments whose volume changes

A compartment with a **rate rule** is fully handled: it joins the state vector,
and every species inside it gains the dilution term `-[S] * (dV/dt) / V` that
follows from `[S] = n/V`. The main vignette works through an example.

A compartment with an **assignment rule** is only partly handled. The volume
itself is correct — it is recomputed in the right-hand side like any other
assignment rule — but the dilution term cannot be formed, because that needs
`dV/dt`, i.e. a symbolic time derivative of the assignment expression, chain
rule and all. Every writer warns:

```{r assignment-compartment}
assigned <- tempfile(fileext = ".xml")
writeLines('<?xml version="1.0" encoding="UTF-8" ?>
<sbml xmlns="http://www.sbml.org/sbml/level3/version2/core" level="3" version="2">
<model substanceUnits="mole" volumeUnits="litre" timeUnits="second" extentUnits="mole">
<listOfCompartments>
<compartment id="c" size="1" spatialDimensions="3" constant="false"/>
</listOfCompartments>
<listOfSpecies>
<species id="A" compartment="c" initialConcentration="2" boundaryCondition="false"
hasOnlySubstanceUnits="false" constant="false"/>
</listOfSpecies>
<listOfParameters>
<parameter id="g" value="1" constant="true"/>
</listOfParameters>
<listOfRules>
<assignmentRule variable="c">
<math xmlns="http://www.w3.org/1998/Math/MathML">
<apply><plus/><cn>1</cn><ci>g</ci></apply></math>
</assignmentRule>
</listOfRules>
</model>
</sbml>', assigned)

tryCatch(
  invisible(capture.output(convertReactions(assigned, tempfile(), format = "R"))),
  warning = function(w) cat(conditionMessage(w), "\n")
)
```

Events that change a compartment size are not handled either, for the same
reason events in general are not.

# Mathematics with no spelling in the target language

None of the following is exercised by the example models, so all of it is
reasoned-correct rather than tested. All of it is **silent**.

**Two-argument logarithms.** The SBML Level 3 writer emits `log(b, x)` meaning
"log of `x` to base `b`". That is correct Julia. It is *reversed* in R, where
`log(x, base)` takes the value first, and invalid in MATLAB, which has no
two-argument `log`. Only the ubiquity writer rewrites it, as a ratio of
`SIMINT_LOGN` calls. A model using `log(b, x)` will produce R code that runs and
returns wrong numbers.

**Other constructs.** `root(n, x)`, `piecewise(...)`, `exponentiale`,
`factorial` in the C++ target, and the `%` modulo operator, whose spelling
differs across R, C++ and MATLAB.

**Logical and relational operators in MATLAB.** `formulaToInfixMatlab()` rewrites
`!` to `~` and `!=` to `~=`. No example model contains a logical operator, so
that rewrite has never actually run against a real model.

**ubiquity's unsupported list.** `ubiquityUnsupported()` reports the constructs
named in the ubiquity documentation — `root`, `piecewise`, `factorial`, `delay`,
`or`, `not`, `xor`. Anything outside that list, the trigonometric functions in
particular, falls into a generic named-call path: it is emitted as a plain call
and reported by name if it cannot be resolved, rather than being caught up
front.

Why this matters more for ubiquity than elsewhere: `build_system()` accepts a
file containing an unknown call, and the failure surfaces much later as a C
compile error naming a shared object, with nothing pointing back at the model.
Hence the `# WARNING` block written into the file itself as well as the R
warning.

# No validation is performed

`getModel()` stops if libSBML reports a problem while *reading* the file, but
reading is not validation. Neither `getModel()` nor `convertReactions()` runs a
consistency check, so a model that loads cleanly may still be semantically
invalid — inconsistent units, an overdetermined system, a species referenced
before it is defined.

If you need validation, run it through libSBML's own validator or the
[online SBML validator](https://sbml.org/validator/) first. `r2sbml` assumes
the model it is given is a correct one.

# Quirks of the query functions

These are not code-generation issues; they affect the inspection API.

## Initial values are reported as the file states them

`getSpeciesIC()` returns whichever of `initialAmount` or `initialConcentration`
the model actually sets — a species carries one or the other, never both — and
names the result by species id:

```{r species-ic}
model <- getModel(ex("sbmlsimple.xml"))
getSpeciesIC(model)
```

**No unit conversion is applied.** That is deliberate, and it is the one thing
here worth watching: a query function should report the file. So a model that
mixes amount-valued and concentration-valued species yields a vector that mixes
units, and `getSpeciesIC()` deliberately disagrees with the initial conditions
`convertReactions()` writes, which are divided by compartment volume because
the generated ODE integrates concentrations:

```{r species-ic-vs-converted}
out <- tempfile(fileext = ".R")
invisible(capture.output(convertReactions(ex("sbmlsimple.xml"), out, format = "R")))
grep("^ +E = [0-9]", readLines(out), value = TRUE)   # 5e-07 = 5e-21 / 1e-14
```

`getSpeciesTable()` shows which column each species used, so you can tell the
units apart:

```{r species-table}
getSpeciesTable(model)[, c("ID", "InitialConcentration", "InitialAmount")]
```

A species that sets neither attribute — its value comes from an initial
assignment or a rule — is reported as `NA` rather than a fabricated `0`. Note
that `NA` there is distinct from the `NaN` in the table above, which marks the
attribute the model simply did not use.

## Names fall back to the id

SBML distinguishes `id`, which every element must have, from `name`, an
optional human label. `getSpeciesNames()` and `getCmtNames()` both return the
`name` where one is set and the `id` otherwise, which is how SBML tools
conventionally display an element:

```{r species-names}
getSpeciesNames(model)
getCmtNames(model)
```

None of the ten examples sets a `name`, so both come back as ids. Use
`getSpeciesTable()`, which keeps `ID` and `Name` in separate columns, if you
need to know whether a label was actually present.

## Absent components raise an error

```{r stopping, error = TRUE}
getRuleMath(model)     # sbmlsimple.xml has no rules
```

This is by design: a query for a component the model does not have is an error,
not a zero-row result. The message names the component, so a caller catching it
can tell "no rules" from "no parameters" without capturing stdout.

# Things that are correct but easy to misread

**`NaN` in `getSpeciesTable()` is not a bug.** It marks the attribute the model
did not set.

**Console output is not part of the return value.** `getRuleMath()` and friends
return `0` invisibly and print their content, so `x <- getRuleMath(m)` gives you
`0`. Use `capture.output()`.

**Compartment volumes are constants in the generated code** — when the model
says they are constant. That is not the volume being ignored.

**ubiquity bounds are `-inf inf`.** Deliberate: an SBML value may be zero or
negative, and a lower bound above the value would be inconsistent. They matter
only for parameter estimation.

**The generated header comment counts model components**, including ones that
were expanded away (function definitions, initial assignments) or not exported
(events). It describes the SBML file, not the generated code.

# Generated code is a starting point

Even where everything above is satisfied, treat the output as a first draft:

- The time span is a placeholder — `[0 10]` for MATLAB, `(0.0, 10.0)` for Julia,
  and nothing at all for the R targets, which leave the call to you.
- Solver tolerances are defaults, and no target's generated code sets them.
  Some models are stiff enough to need tuning: `sbmlmutlicompartment.xml` (rate
  constants from 2500 to 25000) integrates cleanly to `t = 1` with `deSolve`'s
  defaults but exhausts `lsoda`'s step budget before `t = 5`, returning early
  with a warning rather than an error.
- No unit checking is performed. `r2sbml` converts initial conditions between
  amount and concentration using compartment volume, but it does not verify that
  the model's declared units are mutually consistent, and it does not convert
  between, say, millilitres and litres.
- Conversion factors are applied by libSBML's converter, not by `r2sbml`, and
  are only as correct as that converter.

# Where these are tracked

The package keeps `todo.md` in its source repository recording each known gap
along with how it was found, so a claim here can be checked against the
reasoning behind it before anyone acts on it. The test suite covers every
limitation that has a defined behaviour — the warnings, the fallbacks, and the
closed-form solutions used to verify the DAE and dilution paths.
