---
title: "Getting started with r2sbml"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting started with r2sbml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_desolve <- requireNamespace("deSolve", quietly = TRUE)
```

`r2sbml` is a thin R interface to [libSBML](https://sbml.org/software/libsbml/),
built with Rcpp. It does two things:

1. **Reads** an [SBML](https://sbml.org) model and lets you query its components
   from R — species, compartments, parameters, reactions, rules, events and
   function definitions.
2. **Exports** the model's mass balances as ready-to-simulate code for six ODE
   targets: deSolve, mrgsolve, rxode2/nlmixr2, MATLAB, Julia and ubiquity.

It does *not* let you modify a model. There is no setter API and no way to write
SBML back out; `r2sbml` reads and exports.

libSBML itself is bundled with the package and compiled from source during
installation, so nothing needs to be installed separately and the build
downloads nothing. See the README for the compiler and header prerequisites.

```{r library}
library(r2sbml)
```

Ten small example models ship with the package and are used throughout this
vignette:

```{r examples}
basename(list.files(system.file("examples", package = "r2sbml"), pattern = "[.]xml$"))
```

# Reading a model

Everything except `convertReactions()` works in two phases. First `getModel()`
reads the file, keeps the underlying `SBMLDocument` alive, and hands back an
external pointer:

```{r getmodel}
sbml_file <- system.file("examples", "sbmlsimple.xml", package = "r2sbml")
model <- getModel(sbml_file)
class(model)
```

`getModel()` prints a short header as a side effect — the file it read, the
number of diagnostics libSBML reported, and the Level/Version of the document.
If there were any, it prints them and stops with `"Stopping!"` rather than
handing back a pointer you cannot trust.

Reading is not validation. libSBML reports parse-level and structural problems
as it reads, but `getModel()` does not run a consistency check, so a file that
loads cleanly can still be semantically questionable — inconsistent units, say.
`r2sbml` never validates a model.

Every query function then takes that pointer as its only argument.

## Functions that return data

```{r tables}
getSpeciesTable(model)

getReactionTable(model)
```

`getSpeciesTable()` reports both `InitialConcentration` and `InitialAmount`
columns. An SBML species carries one or the other, never both, and the unset one
reads back as `NaN` — so the `NaN` column above is telling you this model is
amount-based, not that anything is wrong.

Simpler accessors return plain vectors:

```{r vectors}
getCmtNames(model)
getCmtSizes(model)
getNumSpecies(model)
```

The full set of data-returning functions:

| Function | Returns |
| --- | --- |
| `getSpeciesTable()`, `getParameterTable()`, `getReactionTable()` | a `data.frame` |
| `getSpeciesNames()`, `getCmtNames()` | a character vector |
| `getSpeciesIC()`, `getCmtSizes()` | a numeric vector |
| `getNumSpecies()` | an integer |

Two of these have sharp edges worth knowing about before you rely on them —
`getSpeciesNames()` returns the SBML `name` attribute (usually empty) and
`getSpeciesIC()` reads only `initialConcentration`. Both are covered in the
companion vignette, `vignette("limitations", package = "r2sbml")`.

## Functions that print

A second family writes to the console and returns `0` invisibly. Assigning them
gets you the `0`, not the text, so capture them instead:

```{r printing}
rules_model <- getModel(system.file("examples", "sbmlassignmentrules.xml",
                                    package = "r2sbml"))
getRuleMath(rules_model)
```

```{r capture}
txt <- capture.output(getReactionMath(model))
txt
```

| Function | Prints |
| --- | --- |
| `getRuleMath()` | assignment, rate and algebraic rules |
| `getReactionMath()` | kinetic laws |
| `getFunctionDefinition()` | function definitions |
| `getEventMath()` | event triggers, delays and assignments |
| `printSBML()`, `echoSBML()` | the document itself |

`printSBML()` and `echoSBML()` are the exception to the two-phase rule: they take
a *filename*, not a model pointer.

## Absent components are an error, not an empty result

A query for a component the model does not have raises an error. It does not
return a zero-row data frame or a zero-length vector:

```{r stopping, error = TRUE}
# sbmlsimple.xml has no rules at all
getRuleMath(model)
```

Note where the information is: the explanatory line goes to the console, while
the condition message itself is only `"Stopping!"`. If you are looping over many
models, wrap the call:

```{r looping}
safely <- function(expr) tryCatch(expr, error = function(e) NULL)
files  <- list.files(system.file("examples", package = "r2sbml"),
                     pattern = "[.]xml$", full.names = TRUE)

has_rules <- vapply(files, function(f) {
  !is.null(safely(capture.output(getRuleMath(getModel(f)))))
}, logical(1))
data.frame(model = basename(files), has_rules = unname(has_rules))
```

The same applies to `getParameterTable()` on a model whose parameters are all
local to its kinetic laws — `sbmlsimple.xml` is exactly that case.

# Exporting mass balances

`convertReactions()` is the other half of the package, and the one function that
does not take a model pointer. It reads the SBML file itself and writes a code
file, returning `NULL` invisibly:

```{r convert-basic}
out_r <- tempfile(fileext = ".R")
convertReactions(sbml_file, out_r, format = "R")
```

| `format` | Target |
| --- | --- |
| `"R"` or `"deSolve"` (default) | [deSolve](https://cran.r-project.org/package=deSolve) |
| `"mrgsolve"` | [mrgsolve](https://mrgsolve.org) |
| `"nlmixr2"` or `"rxode"` | [rxode2](https://nlmixr2.github.io/rxode2/) / nlmixr2 |
| `"MATLAB"` or `"matlab"` | a function file for `ode15s` |
| `"Julia"` or `"julia"` | an `ODEProblem` for DifferentialEquations.jl |
| `"ubiquity"` or `"Ubiquity"` | a [ubiquity](https://r.ubiquity.tools) system file |

An unrecognised format is an error, so a typo cannot leave you with an empty
file that looks like a successful conversion:

```{r bad-format, error = TRUE}
convertReactions(sbml_file, tempfile(), format = "fortran")
```

## What happens before anything is written

`convertReactions()` applies four libSBML converters in sequence:

1. `replaceReactions` — every reaction becomes a rate rule
2. `promoteLocalParameters` — parameters local to a kinetic law become global
3. `expandInitialAssignments` — initial assignments are folded into values
4. `expandFunctionDefinitions` — user-defined functions are inlined

By the time a writer runs, the model is in rate-rule form with no local
parameters and no function definitions left. That is why the generated code
never contains a function definition, and why parameters that were buried inside
a kinetic law appear as top-level constants.

## The deSolve output

```{r show-r}
cat(readLines(out_r), sep = "\n")
```

Three things in that file are worth pointing out.

**The state vector is decided deliberately.** It is not simply "one equation per
rule". A species set by an *assignment rule* is computed inside the right-hand
side and given no slot, because its value follows from the rule at every time
point. A species with *no* rate rule — a boundary condition, or one in no
reaction — keeps its slot with a zero derivative so it still appears in the
solution.

**Initial conditions are converted into the units the ODE integrates.** An SBML
species symbol is a concentration unless `hasOnlySubstanceUnits` is set, and the
rate rules are divided by the compartment volume to match. `sbmlsimple.xml`
declares `initialAmount = 5e-21` in a `1e-14` compartment, so the generated
initial condition is `5e-07`, not `5e-21`. The volume used is the species' own
compartment's, so a multi-compartment model rescales species individually.

**Compartment volumes are emitted as plain constants** when they are constant.
If a compartment carries a rate rule instead, it joins the state vector and is
integrated — see "Changing compartment volumes" below.

You can run the result directly:

```{r run-r, eval = has_desolve}
library(deSolve)

env <- new.env()
source(out_r, local = env)

solution <- ode(y     = env$InitialAmounts,
                times = seq(0, 5, by = 1),
                func  = env$massBalances,
                parms = env$parameters)
round(solution, 9)
```

```{r conservation, eval = has_desolve}
# free + bound enzyme is conserved, and so is total substrate
d <- as.data.frame(solution)
range(d$E + d$ES)
range(d$S + d$ES + d$P)
```

## The other five targets

Each target gets its own spelling of the same mathematics.

```{r mrgsolve}
out_mrg <- tempfile(fileext = ".cpp")
convertReactions(sbml_file, out_mrg, format = "mrgsolve")
cat(readLines(out_mrg), sep = "\n")
```

mrgsolve model blocks are C++, where `^` is bitwise XOR and does not compile for
doubles, so powers are written `pow(a, b)` for this target alone.

```{r rxode}
out_rx <- tempfile(fileext = ".R")
convertReactions(sbml_file, out_rx, format = "nlmixr2")
cat(readLines(out_rx), sep = "\n")
```

```{r matlab}
out_m <- tempfile(fileext = ".m")
convertReactions(sbml_file, out_m, format = "MATLAB")
cat(readLines(out_m), sep = "\n")
```

MATLAB resolves a function by file name, so the generated function is named
after the output path you gave. **Renaming the `.m` file breaks it** — regenerate
instead of renaming.

```{r julia}
out_jl <- tempfile(fileext = ".jl")
convertReactions(sbml_file, out_jl, format = "Julia")
cat(readLines(out_jl), sep = "\n")
```

```{r ubiquity}
out_ub <- tempfile(fileext = ".txt")
convertReactions(sbml_file, out_ub, format = "ubiquity")
cat(readLines(out_ub), sep = "\n")
```

ubiquity has no compartment concept of its own, so a compartment volume becomes
an ordinary `<P>` constant that the rate expressions divide by. It also spells
exponentiation, the transcendental functions and every comparison as bracketed
prefix calls — `SIMINT_POWER[a][b]`, `SIMINT_LOGN[a]`, `SIMINT_GT[a][b]` — which
is why its output looks so different from the rest.

Parameter bounds are written `-inf inf` rather than the `eps inf` you may have
seen in hand-written ubiquity systems: an SBML value may legitimately be zero or
negative, and a lower bound above the value would be inconsistent. They matter
only for estimation.

# Model features that change the shape of the output

## Assignment rules

A species driven by an assignment rule is not integrated. It is recomputed
inside the right-hand side, before the derivatives that read it:

```{r assignment}
assign_file <- system.file("examples", "sbmlassignmentrules.xml", package = "r2sbml")
out_assign  <- tempfile(fileext = ".R")
convertReactions(assign_file, out_assign, format = "R")

lines <- readLines(out_assign)
cat(lines[grep("Assignment Rules", lines):length(lines)], sep = "\n")
```

`S1` and `S2` have no slot in the state vector and no derivative; they are
locals. Giving them slots would leave the solver carrying copies that never
update while the rules recompute different values beside them.

## Algebraic rules make the model a DAE

An algebraic rule is a constraint, `0 = f(...)`, not a derivative. Where the
target can solve a differential-algebraic system, `r2sbml` emits one.

For deSolve that changes the shape of the generated function entirely: it
becomes a **residual** function for `daspk` of `(time, states, derivs, params)`,
not a derivative function for `ode`:

```{r algebraic}
alg_file <- system.file("examples", "sbmlalgebraicrules.xml", package = "r2sbml")
out_alg  <- tempfile(fileext = ".R")
convertReactions(alg_file, out_alg, format = "R")

lines <- readLines(out_alg)
cat(lines[grep("Residuals:", lines)[1]:(grep("return\\(list\\(Residuals", lines))], sep = "\n")
```

The file also defines `InitialDerivatives`, because `daspk` needs `y'` at
`t = 0` consistent with the residuals:

```{r algebraic-solve, eval = has_desolve}
env <- new.env()
source(out_alg, local = env)

sol <- deSolve::daspk(y     = env$InitialAmounts,
                      dy    = env$InitialDerivatives,
                      times = seq(0, 10, by = 2),
                      res   = env$massBalances,
                      parms = env$parameters)
round(sol, 6)
```

```{r algebraic-check, eval = has_desolve}
d <- as.data.frame(sol)
# both algebraic rules hold along the whole trajectory
max(abs(d$E + d$ES - d$E_total))
max(abs(1 * d$E * d$S - (0.5 + 0.5) * d$ES))
```

MATLAB and Julia keep the ordinary derivative shape and use a **singular mass
matrix** instead: `M * y' = f`, with a zero row wherever a constraint sits, so
that row reads `0 = f_i`.

```{r algebraic-matlab}
out_alg_m <- tempfile(fileext = ".m")
convertReactions(alg_file, out_alg_m, format = "MATLAB")
lines <- readLines(out_alg_m)
cat(grep("M = eye|M\\(|MassSingular|ode15s", lines, value = TRUE), sep = "\n")
```

Julia needs both the mass matrix *and* a solver that accepts one, so the
generated header names `Rodas5()` rather than a bare `solve(prob)`:

```{r algebraic-julia}
out_alg_jl <- tempfile(fileext = ".jl")
convertReactions(alg_file, out_alg_jl, format = "Julia")
lines <- readLines(out_alg_jl)
cat(grep("Solve with|^M |^M\\[|mass_matrix|ODEProblem", lines, value = TRUE), sep = "\n")
```

mrgsolve, rxode2 and ubiquity integrate ODEs only. For those, the rule is
written out as a comment and `convertReactions()` warns — see the limitations
vignette.

## Changing compartment volumes

If a compartment carries a rate rule, its volume is integrated alongside the
species, and every species inside it picks up a dilution term. Since a species
symbol is a concentration, `[S] = n/V`, the derivative is

$$\frac{d[S]}{dt} \;=\; \frac{1}{V}\frac{dn}{dt} \;-\; \frac{[S]}{V}\frac{dV}{dt}$$

and libSBML's converter supplies only the first term. `r2sbml` adds the second.

The example models all have constant compartments, so here is a small model
written out by hand — a species sitting in a compartment that grows at a
constant rate, with no reactions at all:

```{r growing-model}
growing <- tempfile(fileext = ".xml")
writeLines('<?xml version="1.0" encoding="UTF-8" ?>
<sbml xmlns="http://www.sbml.org/sbml/level3/version2/core" level="3" version="2">
<model substanceUnits="mole" volumeUnits="litre" timeUnits="second" extentUnits="mole">
<listOfCompartments>
<compartment id="c" size="1" spatialDimensions="3" constant="false"/>
</listOfCompartments>
<listOfSpecies>
<species id="A" compartment="c" initialConcentration="2" boundaryCondition="false"
hasOnlySubstanceUnits="false" constant="false"/>
</listOfSpecies>
<listOfParameters>
<parameter id="g" value="1" constant="true"/>
</listOfParameters>
<listOfRules>
<rateRule variable="c">
<math xmlns="http://www.w3.org/1998/Math/MathML"><ci>g</ci></math>
</rateRule>
</listOfRules>
</model>
</sbml>', growing)

out_grow <- tempfile(fileext = ".R")
convertReactions(growing, out_grow, format = "R")
lines <- readLines(out_grow)
cat(grep("InitialAmounts|^ +[Ac] =|_dt|states\\[", lines, value = TRUE), sep = "\n")
```

`c` is a state, not a constant, and `A` has a non-zero derivative despite taking
part in no reaction — its concentration falls because its container is growing.
The exact solution is `[A](t) = 2 / (1 + t)`, with the amount `A * c` fixed at 2:

```{r growing-solve, eval = has_desolve}
env <- new.env()
source(out_grow, local = env)

times <- c(0, 1, 3, 7)
d <- as.data.frame(deSolve::ode(y = env$InitialAmounts, times = times,
                                func = env$massBalances, parms = env$parameters,
                                rtol = 1e-10, atol = 1e-10))
data.frame(time = times, A = d$A, exact = 2 / (1 + times), amount = d$A * d$c)
```

# Practical notes

**Write to `tempfile()` in examples and tests.** `convertReactions()` overwrites
its output path without asking.

**Expect console output.** `convertReactions()` prints `Conversion completed.`
and the number of ODEs; `getModel()` prints its header. Wrap calls in
`invisible(capture.output(...))` if that noise is unwelcome.

**Watch for warnings.** Several model features cannot be expressed in every
target. Where that is true, `convertReactions()` raises an R warning rather than
silently emitting code that runs but computes the wrong thing. The next vignette
lists every such case, and — just as importantly — the cases where no warning is
raised.

# Where to go next

```r
vignette("limitations", package = "r2sbml")
```

covers what `r2sbml` does not do: unsupported SBML constructs, the per-target
capability matrix, which limitations are announced and which are silent, and the
known quirks of the query functions.
